\pkg{partykit}: A Modular Toolkit for Recursive Partytioning in \proglang{R}

### 2014-03-25 pkg/partykit/inst/mloss/mloss2.Rnw
first draft of user interface section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@551 255f4249-2263-4417-b1ea-8498779e27a7

two blocks: basic infrastructure (next section) and methods for learning trees,
either directly or via coercion.

formula interface

as.party

PMML

print, plot, predict

nodeapply, nodeprune

@@

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\centering
<<mob-plot, echo=FALSE, fig=TRUE, height=7, width=18>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
plot(ttnc_rp, newpage = FALSE, pop = TRUE,
  tp_args = list(reverse = FALSE, fill = grey(c(0.45, 0.85))))
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
plot(ttnc_mob, newpage = FALSE, pop = TRUE, tp_args = list(ylines = 1.3,
  margins = c(1.5, 1, 1.5, 2.2), gp = gpar(fill = grey(c(0.45, 0.85))),
  linecol = "black"))
popViewport(2)
@
\caption{\label{fig} Tree visualizations of survival on Titanic:
\class{rpart} tree converted with \fct{as.party} and visualized by \pkg{partykit} (left); and
logistic-regression-based tree fitted by \fct{glmtree} (right).}
\end{figure}

\section{Internal Structure}

>>>>

The \pkg{partykit} package provides functionality as different levels. First,
there is basic infrastructure for representing, modifying, and displaying trees
and recursive partitions and these tools are mostly intended for developers
and described in the next section. Second, there are tools for inferring trees
from data or for porting trees inferred by other software into \pkg{partykit}.

@@

While originally an important goal for the development of \pkg{partykit} was to
provide infrastructure for the authors' own tree induction algorithms
CTree \citep{Hothorn+Hornik+Zeileis:2006} and MOB \citep{Zeileis+Hothorn+Hornik:2008},
the design was very careful to separate as much functionality as possible into
more general classes that are useful for a far broader class of tree algorithms.
Specifically, to be able to print/plot/predict different trees in a unified way,
there are so-called coercion functions for transforming trees learned in other
software packages (inside and outside of \proglang{R}) into the classes provided
by \pkg{partykit}. Specifically, tree objects learned by \fct{rpart} from the
package of the same name (\citealp{rpart}, implementing CART, \citealp{Breiman+Friedman+Olshen:1984})
and by \fct{J48} from \pkg{RWeka} (\citealp{RWeka}, interfacing \pkg{Weka}'s J4.8
algorithm for C4.5, \citealp{Quinlan:1993}) can be coerced by \fct{as.party}
to the same object class \class{constparty}. This is a general class that can
in principle represent all the major classical tree algorithms with constant
fits in the terminal nodes. Also, the same class is employed for conditional
inference trees (CTree) that can be learned with the \fct{ctree} function
directly within \pkg{partykit} or evolutionary learned trees from \pkg{evtree}
\citep{evtree}.

But not only trees learned within \proglang{R} can be transformed to the
proposed infrastructure but also trees from other software packages. Either
a dedicated interface has to be created using the building blocks described
in the next section (e.g., as done for the J4.8 tree in \pkg{RWeka}) or
PMML (Predictive Model Markup Language) can be used as an intermediate
exchange format. This is an XML standard created by an international
consortium \citep{DMG:2014} that includes a \code{<TreeModel>} tag with
support for constant-fit classification and regression trees. The function
\fct{pmmlTreeModel} allows to read these files and represents them as
\class{simpleparty} objects in \pkg{partykit}. The reason for not using the
\class{constparty} class as above is that the PMML format only stores
point predictions (e.g., a mean or proportion) and not all observations
from the learning sample. So far, the PMML interface has been tested with
output from the \proglang{R} package \pkg{pmml} and \proglang{SPSS}'s
\pkg{AnswerTree} model. The latter includes an implementation of the
QUEST algorithm \citep{Loh+Shih:1997}.

Finally, the \pkg{partykit} function \fct{mob} implements model-based
recursive partitioning (MOB) along with ``mobster'' interfaces for certain
models (e.g., \fct{lmtree}, \fct{glmtree}). These return objects of class
\class{modelparty} where nodes are associated with statistical models (as
opposed to simple constant fits). In principle, this may also be adapted
to other model trees (such as GUIDE, LMT, or M5') but no such interface
is currently available.

All of these different trees (see Table~\ref{tab:overview} for an overview)
use the same infrastructure at the core but possibly with different options
enabled. In all cases, the functions \fct{print}, \fct{plot}, and \fct{predict}
can be used to create textual and graphical displays of the tree and for
computing predictions on new data, respectively. As an example for the
visualizations, Figure~\ref{fig:ttnc} shows two different trees fitted
to the well-known data on survival of passengers on the ill-fated maiden
voyage of the MS~Titanic: The left panel shows a CART tree with constant fits
learned by \pkg{rpart} and converted to \pkg{partykit}. The left panel
shows a MOB tree learned within \pkg{partykit} with a logistic regression
model for treatment effects in the terminal nodes. Additionally, the
are further utility functions, e.g.,
\fct{nodeapply} can be employed to access further information
stored in the nodes of a tree and \fct{nodeprune} can prune selected nodes.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\centering
<<mob-plot, echo=FALSE, fig=TRUE, height=7, width=18>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
plot(ttnc_rp, newpage = FALSE, pop = TRUE,
  tp_args = list(reverse = FALSE, fill = grey(c(0.45, 0.85))))
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
plot(ttnc_mob, newpage = FALSE, pop = TRUE, tp_args = list(ylines = 1.3,
  margins = c(1.5, 1, 1.5, 2.2), gp = gpar(fill = grey(c(0.45, 0.85))),
  linecol = "black"))
popViewport(2)
@
\caption{\label{fig:ttnc} Tree visualizations of survival on Titanic:
\class{rpart} tree converted with \fct{as.party} and visualized by \pkg{partykit} (left); and
logistic-regression-based tree fitted by \fct{glmtree} (right).}
\end{figure}

\section{Developer Infrastructure}
